VariantSpark: Population Scale Clustering of Genotype Information

### 2015-11-27 doc/publications/journal2014/paper.tex
update to paper and letter

\author[
   addressref={aff1,aff2},                   % id's of addresses, e.g. {aff1,aff2}
%   corref={aff1},                       % id of corresponding address, if any
%   noteref={n1},                        % id's of article notes, if any
%   email={jane.e.doe@cambridge.co.uk}   % email address
]{\inits{}\fnm{Aidan R.} \snm{O'Brien}}
\author[
   addressref={aff1},
]{\inits{}\fnm{Neil} \snm{Saunders}}
\author[
   addressref={aff3,aff4},
%   email={john.RS.Smith@cambridge.co.uk}
]{\inits{}\fnm{Fabian A.} \snm{Buske}}
\author[
   addressref={aff1},
   email={Denis.Bauer@CSIRO.au}
%   email={john.RS.Smith@cambridge.co.uk}
]{\inits{}\fnm{Denis C.} \snm{Bauer}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

As shown in Table 1, pre-processing the data is fastest in \variantSpark{}, requiring 3 minutes. %2mins and 58secs. %178s 
This is almost 80\% faster than both the {\sc ADAM} implementation (13 minutes) %(12min and 48secs) %768s
and our Hadoop implementation (14 minutes). %(14min and 22secs). %862s
Unlike \variantSpark{} and our Hadoop implementation, the {\sc ADAM} framework cannot process the VCF files directly but requires them to be converted into a binary {\sc ADAM} file format. 
Although this additional pre-processing step is only required once for each input file, it uses an additional 13 minutes, %13mins and 13secs, %793s
rendering our approach almost an order of magnitude faster (see Figure 1).
R and Python are also slower at pre-processing, taking 67 minutes and 92 minutes, respectively. 
This increased pre-processing time is due to the standard Python and R implementations not natively supporting multithreading, while \variantSpark{}, {\sc ADAM} and Hadoop can use the eight available cores.
The GATK-based VCF to PED conversion for {\sc Admixture} is also slower with over 10 minutes.

@@

%To further demonstrate the scalability of \variantSpark{}, we also cluster the larger 1000 Genomes Project phase 3 VCF files. Although \kMeans{} clustering completes successfully on this dataset,
%it takes over 27 hours (see Table 2) due to more than twice the number of individuals, and hence more variants. For comparison, the uncompressed size of the phase 3 files is 770GB vs the 161GB of the phase 1 dataset.
To further demonstrate the scalability of \variantSpark{}, we also cluster the 1000 Genomes Project phase 3 data, which contains 3000 individuals from 5 super-populations and as a result has over 80 Million variants. 
The uncompressed size of the phase 3 files is 770GB compared to the 161GB of the phase 1 dataset.
\variantSpark{} successfully completes the clustering in just 27 hours (see Table 2) with an ARI of 0.82.

@@

%\variantSpark{} performs clustering on VCF files with over 1000 individuals and 40 million variants in under 15 hours using minimal memory (24GB). 
\variantSpark{} performs clustering on VCF files with over 3000 individuals and 80 million variants in 27 hours using minimal memory (24GB). 
\variantSpark{} supports random genome-wide sampling of variants allowing faster clustering for well characterised cohorts where 20\% of the genome is sufficient. 
On the benchmarking dataset, it outperforms {\sc ADAM} by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
\variantSpark{} utilises {\sc Spark}, which allows for in-memory caching and hence performs 85\% faster than Hadoop (29 min).
\variantSpark{} scales to data sizes that are not feasible to process using R or Python due their requirement to load the whole dataset into memory. 
But even on the small benchmarking dataset, \variantSpark{}'s novel parallelisation approach is faster than traditional multithreading with 94\% speedup over R (75 min), and 95\% over Python (103 min).
\variantSpark{} is also superior in performance and accuracy over the current state-of-the-art tool for individual ancestries determination, {\sc Admixture}.
These benefits of speed, resource consumption and scalability allows \variantSpark{} to be the interface for applying other machine learning tools from MlLib to genomic data. 
Utilising MLlib as well as Spark.ML will enable supervised machine learning applications to e.g. identify variants that jointly interact with phenotypes as well as include electronic health record in addition to the genomic feature vector to e.g. capture medical history as well as predispositions for diagnosis and treatment decisions.

@@

\subsection*{R Implementation}
For our R implementation, we utilize {\sc read.table} to read in a VCF file as a tab-delimited document. As this also captures the first nine columns that don't contain variants, we subsequently remove those columns with {\sc vcf[1:9] <- list()}.
%TODO VariantAnnotation reads in the file quicker but causes issue down the track so overall read.table is faster, right ? - can you add this explanation to the supplementary material.
In the supplementary material we benchmark a slightly faster (by less than one minute) approach of reading the VCF file in using the `VariantAnnotation' package.

@@

\section*{Author's contributions}
    DCB and FAB designed the overall study. ARO implemented \variantSpark\ and conducted the comparison. NS obtained the PGP data and generated the graph.
    All authors contributed to the writing of the document.

\section*{Acknowledgements}
  A.R.O was funded by the NSW Cancer Institute Big Data Big Impact schema, F.A.B by the National Health and Medical Research Council [1051757] and both D.C.B and N.S by Commonwealth Scientific and Industrial Research Organisation's Transformational Capability Platform, Science and Industry Endowment Fund and Information Management and Technology Services. The computation on Azure was funded by Microsoft Azure Research Award. 
The authors would like to thank Piotr Szul and Gareth Williams for their help with setting up Hadoop on the HPC system.

@@

\section*{Additional Files}
  \subsection*{Additional file 1 --- PGP population labels}
    Conversion map self-reported descriptive population labels to super-population labels.

>>>>

\author[
   addressref={aff1,aff2},                   % id's of addresses, e.g. {aff1,aff2}
%   corref={aff1},                       % id of corresponding address, if any
%   noteref={n1},                        % id's of article notes, if any
%   email={jane.e.doe@cambridge.co.uk}   % email address
]{\inits{}\fnm{Aidan R.} \snm{O'Brien}}
\author[
   addressref={aff1},
]{\inits{}\fnm{Neil F.W.} \snm{Saunders}}
\author[
   addressref={aff3,aff4},
%   email={john.RS.Smith@cambridge.co.uk}
]{\inits{}\fnm{Fabian A.} \snm{Buske}}
\author[
   addressref={aff2},
   email={rodney.scott@newcastle.edu.au}
]{\inits{}\fnm{Rodney J.} \snm{Scott}}
\author[
   addressref={aff1},
   email={Denis.Bauer@CSIRO.au}
%   email={john.RS.Smith@cambridge.co.uk}
]{\inits{}\fnm{Denis C.} \snm{Bauer}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors' addresses here        %%
%%                                          %%
%% Repeat \address commands as much as      %%
%% required.                                %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

As shown in Table 1, pre-processing the data is fastest in \variantSpark{}, requiring 3 minutes. %2mins and 58secs. %178s 
This is almost 80\% faster than both the {\sc ADAM} implementation (13 minutes) %(12min and 48secs) %768s
and our Hadoop implementation (14 minutes). %(14min and 22secs). %862s
Unlike \variantSpark{} and our Hadoop implementation, the {\sc ADAM} framework cannot process the VCF files directly but requires them to be converted into a binary {\sc ADAM} file format. 
Although this additional pre-processing step is only required once for each input file, it uses an additional 13 minutes, %13mins and 13secs, %793s
rendering our approach almost an order of magnitude faster (see Figure 1).
R and Python are also slower at pre-processing, taking 34 minutes and 92 minutes, respectively. %34min 30sec % 2070
This increased pre-processing time is due to the standard Python and R implementations not natively supporting multithreading, while \variantSpark{}, {\sc ADAM} and Hadoop can use the eight available cores.
The GATK-based VCF to PED conversion for {\sc Admixture} is also slower with over 10 minutes.

@@

To further demonstrate the scalability of \variantSpark{}, we also cluster the 1000 Genomes Project phase 3 data, which contains 3000 individuals from 5 super-populations and as a result has over 80 Million variants. 
The uncompressed size of the phase 3 files is 770GB compared to the 161GB of the phase 1 dataset.
\variantSpark{} successfully completes the clustering in 27 hours (see Table 2) with an ARI of 0.82.

@@

%TODO add value
\variantSpark{} performs clustering on VCF files with over 3000 individuals and 80 million variants in XX hours using minimal memory (24GB). 
\variantSpark{} supports random genome-wide sampling of variants allowing faster clustering for well characterised cohorts where 20\% of the genome is sufficient. 
On the benchmarking dataset, it outperforms {\sc ADAM} by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
\variantSpark{} utilises {\sc Spark}, which allows for in-memory caching and hence performs 86\% faster than Hadoop (29 min). %7-fold speedup
\variantSpark{} scales to data sizes that are not feasible to process using R or Python due their requirement to load the whole dataset into memory. 
But even on the small benchmarking dataset, \variantSpark{}'s novel parallelisation approach is faster than traditional multithreading with 90\% speedup over R (42 min), and 96\% over Python (103 min). %10 fold %25-fold
\variantSpark{} is also superior in performance and accuracy over the current state-of-the-art tool for individual ancestries determination, {\sc Admixture}.
These benefits of speed, resource consumption and scalability allows \variantSpark{} to be the interface for applying other machine learning tools from MlLib to genomic data. 
Utilising MLlib as well as Spark.ML will enable supervised machine learning applications to e.g. identify variants that jointly interact with phenotypes as well as include electronic health record in addition to the genomic feature vector to e.g. capture medical history as well as predispositions for diagnosis and treatment decisions.

@@

\subsection*{R Implementation}
%For our R implementation, we utilize {\sc read.table} to read in a VCF file as a tab-delimited document. As this also captures the first nine columns that don't contain variants, we subsequently remove those columns with {\sc vcf[1:9] <- list()}.
For our R implementation, we utilize {\sc read.table} to read in a VCF file as a tab-delimited document and remove the first nine columns that do not contain variants. 
%In the supplementary material we benchmark a slightly faster (by less than one minute) approach of reading the VCF file in using the `VariantAnnotation' package.
In the supplementary material we benchmark this against an approach reading the VCF file using the `VariantAnnotation' package, which causes inefficiencies for the subsequent data manipulation tasks and is hence overall slower.

@@

\section*{Author's contributions}
    DCB and FAB designed the overall study. ARO implemented \variantSpark\ and conducted the comparison. NFS obtained the PGP data and generated the graph. RJS guided the interpretation. 
    All authors contributed to the writing of the document.

\section*{Acknowledgements}
  ARO was funded by the NSW Cancer Institute Big Data Big Impact schema, FAB by the National Health and Medical Research Council [1051757] and both DCB and NFS by Commonwealth Scientific and Industrial Research Organisation's Transformational Capability Platform, Science and Industry Endowment Fund and Information Management and Technology Services. The computation on Azure was funded by Microsoft Azure Research Award. 
The authors would like to thank Piotr Szul and Gareth Williams for their help with setting up Hadoop on the HPC system.

@@

\section*{Additional Files}
  \subsection*{Additional file 1 --- PGP population labels and implementation details}
    Conversion map self-reported descriptive population labels to super-population labels. 
    Details of the R implementation.
